Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-02-28 paper/paper.tex
WIP

%We modelled interactions between axon growth cones due to signalling between
%receptors of the same type. For these interactions, repulsion caused a
%movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
%branch $k$ to branch $b$ causing them to move further apart; attraction caused
%the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
%is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_I} \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} > 1.1
\end{equation}
%
$r_{i,k}$ is the expression of receptor $i$ on branch $k$ and the weight here
is defined as a function of the distance, $d_{kb}$, between branches:
%
\begin{equation}
W_{\!_I}(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_{\!_I}}   & d_{kb} \leq 2r_{\!_I} \\
     0 & d_{kb} > 2r_{\!_I}
     \end{cases}
\end{equation}
%
$B_{b,I}$ is the set of growth cones within a distance $2r_{\!_I}$ of branch
$b$. $\hat{\mathbf{x}}_{kb}$ is as defined in Eq.\,\ref{e:C}.

@@

Taken alone, the EphA forward interaction between one branch, $b$, and
another, $k$, will cause a repulsion if $k$ originated more nasally than
$b$. A branch originating on the \emph{temporal} side of $b$ will cause no
repulsion of $b$. However, although $k$ does not induce repulsion in $b$, $b$
should induce a repulsion in $k$ and thus, it could be argued that only a
single ephrin-A ligand gradient should be required in the retina to ensure
that repulsive signals are transmitted between axons whose cell bodies
originte far apart in the retina. However, there is an asymmetry in the
interaction. Consider a cluster of growth cones, $k$, originating from the
nasal retina near to a single cone, $b$, from the temporal retina and assume
that $r_b/r_k>1.1$. The single cone is affected by the sum of 3 interactions
and has a movement of 3 units added to its movement vector (in the temporal
direction). Now reverse the originating locations so that the cluster, $k$,
originates temporally and the individual cone, $b$, comes from the nasal
retina. Now $r_k/r_b>1.1$ and each member of the cluster experiences an
addition of just one unit to its movement vector and the cones move apart at a
speed that is only $1/3$ of that for the original origins.

@@

% Now describe J model, and show the result.

% ./build/sim/agent/agent1 configs/simpler/m_ee_GJ_best_1.json configs/simpler/e_wt_figcomp3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_wt_figcomp3_exit_true_steps_1500.png}
\caption{$\mathbf{GJ}$ model}
\label{f:GJ}
\end{figure}

@@

%% TO ADD: Figure for the Reber/Brown manipulation on the GJ model.

@@

%% HEREish
Next, we investigated the idea that the same receptor-ligand
interaction that leads to gradient-based movement of RGC axons on the tectum
could also act as a signal between the axons. This is plausible, because
retinal ganglion cells are known to express ephrin receptor ligands as well as
Eph receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$,
expressed by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C).
%
We combined chemotaxis with a multiplicative (or mass-action?) interaction,
$\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the movement
vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
 %%+  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines
whether the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$
is the expression of ligand $i$ on branch $k$.

Considering just one receptor-ligand pair, for instance $r_0$/$l_0$, and
defining their interaction as repulsive, an axon from the transverse side of
the retina, which expresses a high level of receptor $r_0$ will interact most
strongly with an axon from the nasal side of the retina, where $l_0$ is most
highly expressed. The interaction with axons that originate nearby, on the
transverse side will be minimal, as these express low levels of $l_0$. Axons
originating from the nasal side of the retina also interact most strongly with
axons originating nearby and in general the $r_0$/$l_0$ interaction is weaker
between a nasal axon and a second retinal axon. However, when the opposing
receptor/ligand pair ($r_2$/$l_2$) interaction is summed with that of
$r_0$/$l_0$, a more even interaction is found, with interactions between axons
from extreme sides of the retina being the strongest. (Poss.~supp.~figure
based on the Octave script J.m).

We performed the same simulated annealing process as for the
$\mathbf{GC}$ model to find the parameters best able to recover the wildtype
pattern and found $m_{\!_G} = 0.00047$, $m_{\!_J} = 0.00089$ and $r_{\!_J} =
0.341$.

See Fig.\,\ref{f:GJ} and Fig.\,\ref{f:GJsurg}.

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{./images/model_vs_noise.png}
\caption{ \textbf{A} Chemotaxis-only model. The mean value of $\eta$ (number of crossings) and $\epsilon$
(mean distance error), averaged over the period from simulation time $t=750$ to $t=1000$, plotted with respect to the
noise gain ($\nu$) in the best $\mathbf{G}$ model and the wildtype condition (compare with
Fig.\,\ref{f:G0}). Noise for $\nu < 0.5$ is tolerated by the model, but
disorder increases about this value.
\textbf{B} Competitive interaction in the $\mathbf{GI}$ model extends the
range of ligand gradient noise which can be tolerated by the system.
Both metrics indicate that a moderate
amount of gradient sampling noise is \emph{beneficial} to the establishment of
an ordered tectum.}
\label{f:noise}
\end{figure}

@@

Fig.\,\ref{f:noise} shows the response of the $\mathbf{GI}$ model to the
addition of ligand gradient estimation noise.

@@

>>>>

%We modelled interactions between axon growth cones due to signalling between
%receptors of the same type. For these interactions, repulsion caused a
%movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
%branch $k$ to branch $b$ causing them to move further apart; attraction caused
%the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
%is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_I} \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} > 1.1
\end{equation}
%
$r_{i,k}$ is the expression of receptor $i$ on branch $k$ and the weight here
is defined as a function of the distance, $d_{kb}$, between branches:
%
\begin{equation} \label{e:WI}
W_{\!_I}(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_{\!_I}}   & d_{kb} \leq 2r_{\!_I} \\
     0 & d_{kb} > 2r_{\!_I}
     \end{cases}
\end{equation}
%
$B_{b,I}$ is the set of growth cones within a distance $2r_{\!_I}$ of branch
$b$. $\hat{\mathbf{x}}_{kb}$ is as defined in Eq.\,\ref{e:C}.

@@

Taken alone, the EphA forward interaction between one RGC branch, $b$, and
another, $k$, will cause a repulsion if $k$ originated more nasally than
$b$. A branch originating on the \emph{temporal} side of $b$ will cause no
repulsion of $b$. However, although $k$ does not induce repulsion in $b$, $b$
should induce a repulsion in $k$ and thus, it could be argued that only a
single ephrin-A ligand gradient should be required in the retina to ensure
that repulsive signals are transmitted between axons whose cell bodies
originte far apart in the retina. However, there is an asymmetry in the
interaction. Consider a cluster of growth cones, $k$, originating from the
nasal retina near to a single cone, $b$, from the temporal retina and assume
that $r_b/r_k>1.1$. The single cone is affected by the sum of 3 interactions
and has a movement of 3 units added to its movement vector (in the temporal
direction). Now reverse the originating locations so that the cluster, $k$,
originates temporally and the individual cone, $b$, comes from the nasal
retina. Now $r_k/r_b>1.1$ and each member of the cluster experiences an
addition of just one unit to its movement vector and the cones move apart at a
speed that is only $1/3$ of that for the original origins.

@@

% Now describe J model, and show the result.
We defined a movement vector arising from a multiplicative receptor-ligand
interaction, $\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the
movement vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
% noting that $m_J J_b$ \emph{replaces} $m_c C_b$.
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
\mathbf{J}_b = \frac{1}{|B_{b,J}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_J}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $W_{\!_J}$ is distance based weighting analogous to $W_{\!_I}$,
$B_{b,J}$ is analogous to $B_{b,I}$ and $Q_{\!_J}(d_{kb})$ is the signalling
strength between two growth cones of receptor-ligand interaction radius
$r_{\!_J}$ a distance $d_{kb}$ from one another, given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
                 F_i & \sum_i^N r_{i,b}\,l_{i,k} > s,~d_{kb} \leq 2r_{\!_J} \\
                 0 & \mathrm{otherwise}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ determines whether the interaction, $\mathbf{J}_b$, is
repulsive or attractive. $l_{i,k}$ is the expression of ligand $i$ on branch
$k$.

$\mathbf{J}_b$ would be equivalent to $\mathbf{I}_b$ if there was a single
$l_i$ expressed in the retina and its form was $1/r_i$. As well as allowing
for two, opp

With $m_{\!_G}$ fixed at the value 0.0012, and allowing the threshold
parameter $s$ to vary along with $m_{\!_J}$ and $r_{\!_J}$ we performed a
simulated annealing optimization. The parameters best able to recover the
wildtype pattern were $s=4.732$, $m_{\!_J} = 0.247$ and $r_{\!_J} =
0.079$. The simulation of the system with these parameters is shown in
Fig.\,\ref{f:GJ} and Fig.\,\ref{f:GJsurg}.

% ./build/sim/agent/agent1 configs/simpler/m_ee_GJ_best_1.json configs/simpler/e_wt_figcomp3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_wt_figcomp3_exit_true_steps_1500.png}
\caption{Chemoaffinity and receptor-ligand mediated competitive interactions
($\mathbf{GJ}$ model). }
\label{f:GJ}
\end{figure}

@@

\subsection*{Knockin and Knockdown manipulations}

@@

%Considering just one receptor-ligand pair, for instance $r_0$/$l_0$, and
%defining their interaction as repulsive, an axon from the transverse side of
%the retina, which expresses a high level of receptor $r_0$ will interact most
%strongly with an axon from the nasal side of the retina, where $l_0$ is most
%highly expressed. The interaction with axons that originate nearby, on the
%transverse side will be minimal, as these express low levels of $l_0$. Axons
%originating from the nasal side of the retina also interact most strongly with
%axons originating nearby and in general the $r_0$/$l_0$ interaction is weaker
%between a nasal axon and a second retinal axon. However, when the opposing
%receptor/ligand pair ($r_2$/$l_2$) interaction is summed with that of
%$r_0$/$l_0$, a more even interaction is found, with interactions between axons
%from extreme sides of the retina being the strongest. (Poss.~supp.~figure
%based on the Octave script J.m).

@@

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{./images/model_vs_noise.png}
\caption{ \textbf{A} Chemotaxis-only model. The mean value of $\eta$ (number of crossings) and $\epsilon$
(mean distance error), averaged over the period from simulation time $t=750$ to $t=1000$, plotted with respect to the
noise gain ($\nu$) in the best $\mathbf{G}$ model and the wildtype condition (compare with
Fig.\,\ref{f:G0}). Noise for $\nu < 0.5$ is tolerated by the model, but
disorder increases above this value.
\textbf{B} Competitive interaction in the $\mathbf{GJ}$ model extends the
range of ligand gradient noise which can be tolerated by the system.
Both metrics indicate that a moderate
amount of gradient sampling noise is \emph{beneficial} to the establishment of
an ordered tectum.}
\label{f:noise}
\end{figure}

@@

Fig.\,\ref{f:noise} shows the response of both the chemoaffinity-only model
and the full $\mathbf{GJ}$ model to the addition of ligand gradient estimation
noise. The results indicate that a moderate amount of noise is beneficial for
the establishment of an ordered map and that the competitive model is much
more robust to higher levels of gradient estimation noise.

@@

Discuss the requirement for a distance based weight and the fact that without
this, the system doesn't converge on a stable solution.
